Plant Trait Imputation

### 2024-06-11 code/L1/L1_P5_TropicalAndes_plant_imputetraits.Rmd
add row counts and length outputs, update trait list, harmonize and clean new traits

---
title: "Plant Trait Imputation"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script fill plant trait gaps with imputation."
data input: "ADD"
data output: "ADD"
date: "2023-10-04"
output: html_document
---

# Set file paths
```{r}
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

# Load required packages
```{r}
library(tidyr)
library(BIEN)
library(GIFT)
library(mice)
library(dplyr)
```

@@

# get a list of species
na_species_list <- unique(plant_traits$species)
```

@@

# retrieve list of species without family & genus information
species_no_family <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
species_no_genus <- na_species_df %>%
  filter(is.na(genus) | genus == "") %>%
  select(species) %>%
  pull()
# check if all the species in no_family & no_genus lists are the same
identical(species_no_family, species_no_genus)
```

@@

# retrieve list of species without family & genus information
species_no_family_2 <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
```

@@

# retrieve list of species without family & genus information
species_no_family_3 <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
```

@@

# Separate dataframes from list
DispersalSyndrome <- trait_dfs$DispersalSyndrome
PlantLifespan_years <- trait_dfs$PlantLifespan_years
FruitMass_mg <- trait_dfs$FruitMass_mg
PlantHeight_m <- trait_dfs$PlantHeight_m
GrowthForm <- trait_dfs$GrowthForm
FruitLength_mm <- trait_dfs$FruitLength_mm
SeedWidth_mm <- trait_dfs$SeedWidth_mm
SeedLength_mm <- trait_dfs$SeedLength_mm
SeedMass_g <- trait_dfs$SeedMass_g
```

## Create a list of unique genus and families for each trait
```{r}
DispersalSyndrome_families <- unique(DispersalSyndrome$family)
DispersalSyndrome_genus <- unique(DispersalSyndrome$genus)
length(DispersalSyndrome_families)
length(DispersalSyndrome_genus)
```

```{r}
PlantLifespan_years_families <- unique(PlantLifespan_years$family)
PlantLifespan_years_genus <- unique(PlantLifespan_years$genus)
length(PlantLifespan_years_families)
length(PlantLifespan_years_genus)
```

```{r}

FruitMass_mg_families <- unique(FruitMass_mg$family)
FruitMass_mg_genus <- unique(FruitMass_mg$genus)

PlantHeight_m_families <- unique(PlantHeight_m$family)
PlantHeight_m_genus <- unique(PlantHeight_m$genus)

GrowthForm_families <- unique(GrowthForm$family)
GrowthForm_genus <- unique(GrowthForm$genus)

FruitLength_mm_families <- unique(FruitLength_mm$family)
FruitLength_mm_genus <- unique(FruitLength_mm$genus)

SeedWidth_mm_families <- unique(SeedWidth_mm$family)
SeedWidth_mm_genus <- unique(SeedWidth_mm$genus)

SeedLength_mm_families <- unique(SeedLength_mm$family)
SeedLength_mm_genus <- unique(SeedLength_mm$genus)

SeedMass_g_families <- unique(SeedMass_g$family)
SeedMass_g_genus <- unique(SeedMass_g$genus)
```

## Retrive trait records for species using BIEN & GIFT
### Dispersal Syndrome
```{r}
DispersalSyndrome_families_BIEN <- BIEN_trait_traitbyfamily(family = DispersalSyndrome_families, trait = "whole plant dispersal syndrome")
DispersalSyndrome_genus_BIEN <- BIEN_trait_traitbygenus(genus = DispersalSyndrome_genus, trait = "whole plant dispersal syndrome")

@@

```{r}
nrow(DispersalSyndrome_families_GIFT)
```

### Fruit Length
```{r}
Fruit.Length_families_BIEN <- BIEN_trait_traitbyfamily(family = Fruit.Length_families, trait = c("maximum fruit length", "minimum fruit length"))
Fruit.Length_genus_BIEN <- BIEN_trait_traitbygenus(genus = Fruit.Length_genus, trait = c("maximum fruit length", "minimum fruit length"))

@@

```{r}
nrow(DispersalSyndrome_families_BIEN)
length(unique(DispersalSyndrome_families_BIEN$scrubbed_family))
unique(DispersalSyndrome_families_BIEN$trait_value)
```

```{r}
# Plant Height
Plant.Height_families_BIEN <- BIEN_trait_traitbyfamily(family = Plant.Height_families, trait = "whole plant height")
Plant.Height_genus_BIEN <- BIEN_trait_traitbygenus(genus = Plant.Height_genus, trait = "whole plant height")

@@

```{r}
# Fruit Type
Fruit.Type_families_BIEN <- BIEN_trait_traitbyfamily(family = Fruit.Type_families, trait = "fruit type")
Fruit.Type_genus_BIEN <- BIEN_trait_traitbygenus(genus = Fruit.Type_genus, trait = "fruit type")

```{r}
# Growth Form
Growth.Form_families_BIEN <- BIEN_trait_traitbyfamily(family = Growth.Form_families, trait = "whole plant growth form")
Growth.Form_genus_BIEN <- BIEN_trait_traitbygenus(genus = Growth.Form_genus, trait = "whole plant growth form")

## Combine new info with existing trait table
Make sure numeric traits match the units of existing trait table. 
To fill in gaps of missing traits, use genus level data first then family level. Add column to trait dataframe of the level of trait measurement: species, genus, family.
```{r}
# for traits to family level from BIEN, keep columns of scrubbed_family, trait_name, trait_value
Dispersal.Syndrome_families_BIEN_df <- Dispersal.Syndrome_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
Fruit.Length_families_BIEN_df <- Fruit.Length_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
Plant.Height_families_BIEN_df <- Plant.Height_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
Fruit.Type_families_BIEN_df <- Fruit.Type_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
Growth.Form_families_BIEN_df <- Growth.Form_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
```

```{r}
# for traits to genus level from BIEN, keep columns of scrubbed_genus, trait_name, trait_value
Dispersal.Syndrome_genus_BIEN_df <- Dispersal.Syndrome_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
Fruit.Length_genus_BIEN_df <- Fruit.Length_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
Plant.Height_genus_BIEN_df <- Plant.Height_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
Fruit.Type_genus_BIEN_df <- Fruit.Type_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
Growth.Form_genus_BIEN_df <- Growth.Form_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
```

Get a single trait value for each family/genus
```{r}
# numeric traits
numeric_traits_combined <- function(df, level) {
  df$trait_value <- as.numeric(df$trait_value)
  if (level == "genus") {
    df %>%
      group_by(scrubbed_genus) %>%
      summarise(trait_value = mean(trait_value, na.rm = TRUE)) %>%
      ungroup()
  } else if (level == "family") {
    df %>%
      group_by(scrubbed_family) %>%
      summarise(trait_value = mean(trait_value, na.rm = TRUE)) %>%
      ungroup()
  }
}

# non-numeric traits
cat_traits_combined <- function(df, level){
  if (level == "genus") {
    df %>%
      group_by(scrubbed_genus) %>%
      summarise_all(.funs = list(mode_trait = ~names(sort(table(.), decreasing = TRUE))[1])) %>%
      ungroup()
  } else if (level == "family") {
    df %>%
      group_by(scrubbed_family) %>%
      summarise_all(.funs = list(mode_trait = ~names(sort(table(.), decreasing = TRUE))[1])) %>%
      ungroup()
  }
}

@@

```{r}
Dispersal.Syndrome_families_BIEN_df <- cat_traits_combined(Dispersal.Syndrome_families_BIEN_df, "family")
Dispersal.Syndrome_genus_BIEN_df <- cat_traits_combined(Dispersal.Syndrome_genus_BIEN_df, "genus")
```

```{r}
Fruit.Length_families_BIEN_df <- numeric_traits_combined(Fruit.Length_families_BIEN_df, "family")
Fruit.Length_genus_BIEN_df <- numeric_traits_combined(Fruit.Length_genus_BIEN_df, "genus")
```

```{r}
Plant.Height_families_BIEN_df <- numeric_traits_combined(Plant.Height_families_BIEN_df, "family")
Plant.Height_genus_BIEN_df <- numeric_traits_combined(Plant.Height_genus_BIEN_df, "genus")
```

```{r}
Fruit.Type_families_BIEN_df <- cat_traits_combined(Fruit.Type_families_BIEN_df, "family")
Fruit.Type_genus_BIEN_df <- cat_traits_combined(Fruit.Type_genus_BIEN_df, "genus")
```

```{r}
Growth.Form_families_BIEN_df <- cat_traits_combined(Growth.Form_families_BIEN_df, "family")
Growth.Form_genus_BIEN_df <- cat_traits_combined(Growth.Form_genus_BIEN_df, "genus")
```

add column of trait level (species, genus, family)
```{r}
# Add new column trait_level to long_plant_trait dataframe
long_plant_traits <- long_plant_traits %>%
  mutate(TraitLevel = ifelse(!is.na(TraitValue), "species", NA))
```

>>>>

---
title: "Plant Trait Imputation"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script fill plant trait gaps with imputation."
data input: "TropicalAndes_all_plant_traits_standardized.csv"
data output: "TropicalAndes_imputed_plant_traits.csv"
date: "2023-10-04"
output: html_document
---

# Set file paths
```{r}
data_path_L1 <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

# Load required packages
```{r}
library(tidyr)
library(BIEN)
library(GIFT)
library(purrr)
library(mice)
library(dplyr)
```

@@

# Convert the wide dataframe to long format
long_plant_traits <- plant_traits %>%
  gather(TraitName, TraitValue, -species)
# Filter rows with NA values
na_traits <- long_plant_traits %>%
  filter(is.na(TraitValue))
dim(na_traits)

# get a list of species
na_species_list <- unique(plant_traits$species)
length(na_species_list)
```

@@

# retrieve list of species without family & genus information
species_no_family <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
species_no_genus <- na_species_df %>%
  filter(is.na(genus) | genus == "") %>%
  select(species) %>%
  pull()

# check if all the species in no_family & no_genus lists are the same
identical(species_no_family, species_no_genus)

length(species_no_family)
```

@@

# retrieve list of species without family & genus information
species_no_family_2 <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
length(species_no_family_2)
```

@@

# retrieve list of species without family & genus information
species_no_family_3 <- na_species_df %>%
  filter(is.na(family) | family == "") %>%
  select(species) %>%
  pull()
length(species_no_family_3)
```

@@

# Separate dataframes from list
PlantLifespan_years <- trait_dfs$PlantLifespan_years
FruitType <- trait_dfs$FruitType
DispersalSyndrome <- trait_dfs$DispersalSyndrome
FruitMass_mg <- trait_dfs$FruitMass_mg
SeedLength_mm <- trait_dfs$SeedLength_mm
SeedWidth_mm <- trait_dfs$SeedWidth_mm
GrowthForm <- trait_dfs$GrowthForm
FruitLength_mm <- trait_dfs$FruitLength_mm
SeedMass_g <- trait_dfs$SeedMass_g
PlantHeight_m <- trait_dfs$PlantHeight_m
```

## Create a list of unique genus and families for each trait
```{r}
PlantLifespan_years_families <- unique(PlantLifespan_years$family)
PlantLifespan_years_genus <- unique(PlantLifespan_years$genus)
length(PlantLifespan_years_families)
length(PlantLifespan_years_genus)
```

```{r}
FruitType_families <- unique(FruitType$family)
FruitType_genus <- unique(FruitType$genus)
length(FruitType_families)
length(FruitType_genus)
```

```{r}
DispersalSyndrome_families <- unique(DispersalSyndrome$family)
DispersalSyndrome_genus <- unique(DispersalSyndrome$genus)
length(DispersalSyndrome_families)
length(DispersalSyndrome_genus)
```

```{r}
FruitMass_mg_families <- unique(FruitMass_mg$family)
FruitMass_mg_genus <- unique(FruitMass_mg$genus)
length(FruitMass_mg_families)
length(FruitMass_mg_genus)
```

```{r}
SeedLength_mm_families <- unique(SeedLength_mm$family)
SeedLength_mm_genus <- unique(SeedLength_mm$genus)
length(SeedLength_mm_families)
length(SeedLength_mm_genus)
```

```{r}
SeedWidth_mm_families <- unique(SeedWidth_mm$family)
SeedWidth_mm_genus <- unique(SeedWidth_mm$genus)
length(SeedWidth_mm_families)
length(SeedWidth_mm_genus)
```

```{r}
GrowthForm_families <- unique(GrowthForm$family)
GrowthForm_genus <- unique(GrowthForm$genus)
length(GrowthForm_families)
length(GrowthForm_genus)
```

```{r}
FruitLength_mm_families <- unique(FruitLength_mm$family)
FruitLength_mm_genus <- unique(FruitLength_mm$genus)
length(FruitLength_mm_families)
length(FruitLength_mm_genus)
```

```{r}
SeedMass_g_families <- unique(SeedMass_g$family)
SeedMass_g_genus <- unique(SeedMass_g$genus)
length(SeedMass_g_families)
length(SeedMass_g_genus)
```

```{r}
PlantHeight_m_families <- unique(PlantHeight_m$family)
PlantHeight_m_genus <- unique(PlantHeight_m$genus)
length(PlantHeight_m_families)
length(PlantHeight_m_genus)
```

## Retrive trait records for species using BIEN & GIFT
### Plant Lifespan
```{r}
PlantLifespan_years_families_BIEN <- BIEN_trait_traitbyfamily(family = PlantLifespan_years_families, trait = "maximum whole plant longevity")
PlantLifespan_years_genus_BIEN <- BIEN_trait_traitbygenus(genus = PlantLifespan_years_genus, trait = "maximum whole plant longevity")

# Error in GIFT_traits_tax(trait_IDs = "2.2.1") :   None of the traits asked was available at the taxonomic level.
#PlantLifespan_years_families_GIFT <- GIFT_traits_tax(trait_IDs = "2.2.1")
#PlantLifespan_years_families_GIFT <- PlantLifespan_years_families_GIFT[PlantLifespan_years_families_GIFT$family %in% PlantLifespan_years_families, ]
```

```{r}
nrow(PlantLifespan_years_families_BIEN)
length(unique(PlantLifespan_years_families_BIEN$scrubbed_family))
unique(PlantLifespan_years_families_BIEN$trait_value)
```

```{r}
nrow(PlantLifespan_years_genus_BIEN)
length(unique(PlantLifespan_years_genus_BIEN$scrubbed_genus))
unique(PlantLifespan_years_genus_BIEN$trait_value)
```

### Fruit Type
```{r}
FruitType_families_BIEN <- BIEN_trait_traitbyfamily(family = FruitType_families, trait = "fruit type")
FruitType_genus_BIEN <- BIEN_trait_traitbygenus(genus = FruitType_genus, trait = "fruit type")

```{r}
nrow(FruitType_families_BIEN)
length(unique(FruitType_families_BIEN$scrubbed_family))
unique(FruitType_families_BIEN$trait_value)
```

```{r}
nrow(FruitType_genus_BIEN)
length(unique(FruitType_genus_BIEN$scrubbed_genus))
unique(FruitType_genus_BIEN$trait_value)
```

### Dispersal Syndrome
```{r}
DispersalSyndrome_families_BIEN <- BIEN_trait_traitbyfamily(family = DispersalSyndrome_families, trait = "whole plant dispersal syndrome")
DispersalSyndrome_genus_BIEN <- BIEN_trait_traitbygenus(genus = DispersalSyndrome_genus, trait = "whole plant dispersal syndrome")

@@

```{r}
nrow(DispersalSyndrome_families_GIFT)
```
This information is not useful to fill in trait gaps.

### Fruit Mass
No information in BIEN or GIFT

### Seed Length
Only in GIFT: Seed_length_min (3.10.1), Seed_length_max (3.10.2), Seed_length_mean (3.10.3)
```{r}
SeedLength_mm_families_GIFT <- GIFT_traits_tax(trait_IDs = c("3.10.1", "3.10.2", "3.10.3"))
SeedLength_mm_families_GIFT <- SeedLength_mm_families_GIFT[SeedLength_mm_families_GIFT$family %in% SeedLength_mm_families, ]
```

```{r}
nrow(SeedLength_mm_families_GIFT)
```

### Seed Width
Only in GIFT: Seed_width_min (3.11.1), Seed_width_max (3.11.2), Seed_width_mean (3.11.3)
```{r}
#Error in GIFT_traits_tax(trait_IDs = c("3.11.1", "3.11.2", "3.11.3")) :   None of the traits asked was available at the taxonomic level.
#SeedWidth_mm_families_GIFT <- GIFT_traits_tax(trait_IDs = c("3.11.1", "3.11.2", "3.11.3"))
#SeedWidth_mm_families_GIFT <- SeedWidth_mm_families_GIFT[SeedWidth_mm_families_GIFT$family %in% SeedWidth_mm_families, ]
```

### Growth Form
```{r}
# Growth Form
GrowthForm_families_BIEN <- BIEN_trait_traitbyfamily(family = GrowthForm_families, trait = "whole plant growth form")
GrowthForm_genus_BIEN <- BIEN_trait_traitbygenus(genus = GrowthForm_genus, trait = "whole plant growth form")

```{r}
nrow(GrowthForm_families_BIEN)
length(unique(GrowthForm_families_BIEN$scrubbed_family))
unique(GrowthForm_families_BIEN$trait_value)
```

```{r}
nrow(GrowthForm_genus_BIEN)
length(unique(GrowthForm_genus_BIEN$scrubbed_genus))
unique(GrowthForm_genus_BIEN$trait_value)
```

### Fruit Length
```{r}
FruitLength_mm_families_BIEN <- BIEN_trait_traitbyfamily(family = FruitLength_mm_families, trait = c("maximum fruit length", "minimum fruit length"))
FruitLength_mm_genus_BIEN <- BIEN_trait_traitbygenus(genus = FruitLength_mm_genus, trait = c("maximum fruit length", "minimum fruit length"))

@@

```{r}
nrow(FruitLength_mm_families_BIEN)
length(unique(FruitLength_mm_families_BIEN$scrubbed_family))
unique(FruitLength_mm_families_BIEN$trait_value)
```

```{r}
nrow(FruitLength_mm_genus_BIEN)
length(unique(FruitLength_mm_genus_BIEN$scrubbed_genus))
unique(FruitLength_mm_genus_BIEN$trait_value)
```

### Seed Mass
```{r}
SeedMass_g_families_BIEN <- BIEN_trait_traitbyfamily(family = SeedMass_g_families, trait = "seed mass")
SeedMass_g_genus_BIEN <- BIEN_trait_traitbygenus(genus = SeedMass_g_genus, trait = "seed mass")

SeedMass_g_families_GIFT <- GIFT_traits_tax(trait_IDs = c("3.2.1", "3.2.2", "3.2.3"))
SeedMass_g_families_GIFT <- SeedMass_g_families_GIFT[SeedMass_g_families_GIFT$family %in% SeedMass_g_families, ]
```

```{r}
nrow(SeedMass_g_families_BIEN)
length(unique(SeedMass_g_families_BIEN$scrubbed_family))
unique(SeedMass_g_families_BIEN$trait_value)
```

```{r}
nrow(SeedMass_g_genus_BIEN)
length(unique(SeedMass_g_genus_BIEN$scrubbed_genus))
unique(SeedMass_g_genus_BIEN$trait_value)
```

```{r}
nrow(SeedMass_g_families_GIFT)
```

### Plant Height
```{r}
PlantHeight_m_families_BIEN <- BIEN_trait_traitbyfamily(family = PlantHeight_m_families, trait = "whole plant height")
PlantHeight_m_genus_BIEN <- BIEN_trait_traitbygenus(genus = PlantHeight_m_genus, trait = "whole plant height")

@@

```{r}
nrow(PlantHeight_m_families_BIEN)
length(unique(PlantHeight_m_families_BIEN$scrubbed_family))
unique(PlantHeight_m_families_BIEN$trait_value)
```

```{r}
nrow(PlantHeight_m_genus_BIEN)
length(unique(PlantHeight_m_genus_BIEN$scrubbed_genus))
unique(PlantHeight_m_genus_BIEN$trait_value)
```

## Clean new trait data
Make sure numeric traits match the units of existing trait table. Numeric traits should only contain numeric values and non-numeric traits should only have non-numeric trait values. 
### # Lifespan
remove non-numeric trait values
```{r}
PlantLifespan_years_families_BIEN <- PlantLifespan_years_families_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(PlantLifespan_years_families_BIEN)
unique(PlantLifespan_years_families_BIEN$unit)

PlantLifespan_years_genus_BIEN <- PlantLifespan_years_genus_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(PlantLifespan_years_genus_BIEN)
unique(PlantLifespan_years_genus_BIEN$unit)
```

# Fruit type
haromonize trait values
```{r}
unique(FruitType_families_BIEN$trait_value)
unique(FruitType_genus_BIEN$trait_value)
FruitType_families_BIEN <- FruitType_families_BIEN %>%
  filter() %>%
  mutate(trait_value = case_when(
    grepl('Berry|berry|Berry ', trait_value, ignore.case = TRUE) ~ "berry",
    grepl('capsule|Capsule|Capsule ', trait_value, ignore.case = TRUE) ~ "capsule",
    grepl('drupe|Drupe|Drupe |Drupaceous', trait_value, ignore.case = TRUE) ~ "drupe",
    grepl('Squash', trait_value, ignore.case = TRUE) ~ "squash",
    grepl('pod|Pod|Pod ', trait_value, ignore.case = TRUE) ~ "pod",
    grepl('Samara|Samaroid', trait_value, ignore.case = TRUE) ~ "samara"
  )) %>%
  filter(!is.na(trait_value))
FruitType_genus_BIEN <- FruitType_genus_BIEN %>%
  filter() %>%
  mutate(trait_value = case_when(
    grepl('Berry|berry|Berry ', trait_value, ignore.case = TRUE) ~ "berry",
    grepl('capsule|Capsule|Capsule ', trait_value, ignore.case = TRUE) ~ "capsule",
    grepl('drupe|Drupe|Drupe |Drupaceous', trait_value, ignore.case = TRUE) ~ "drupe",
    grepl('Squash', trait_value, ignore.case = TRUE) ~ "squash",
    grepl('pod|Pod|Pod ', trait_value, ignore.case = TRUE) ~ "pod",
    grepl('Samara|Samaroid', trait_value, ignore.case = TRUE) ~ "samara"
  )) %>%
  filter(!is.na(trait_value))
unique(FruitType_families_BIEN$trait_value)
unique(FruitType_genus_BIEN$trait_value)
```

# Growth form
harmonize trait values and remove numeric values
```{r}
# remove numeric rows
GrowthForm_families_BIEN <- GrowthForm_families_BIEN %>%
  filter(is.na(as.numeric(trait_value)))
nrow(GrowthForm_families_BIEN)
GrowthForm_genus_BIEN <- GrowthForm_genus_BIEN %>%
  filter(is.na(as.numeric(trait_value)))
nrow(GrowthForm_genus_BIEN)
```

```{r}
# classify values as herb, shrub, tree, or other
GrowthForm_families_BIEN <- GrowthForm_families_BIEN %>%
  mutate(Category = case_when(
    grepl("herb", trait_value, ignore.case = TRUE) ~ "herb",
    grepl("shrub", trait_value, ignore.case = TRUE) ~ "shrub",
    grepl("tree", trait_value, ignore.case = TRUE) ~ "tree",
    TRUE ~ "other"
  ))
GrowthForm_genus_BIEN <- GrowthForm_genus_BIEN %>%
  mutate(Category = case_when(
    grepl("herb", trait_value, ignore.case = TRUE) ~ "herb",
    grepl("shrub", trait_value, ignore.case = TRUE) ~ "shrub",
    grepl("tree", trait_value, ignore.case = TRUE) ~ "tree",
    TRUE ~ "other"
  ))
```

### Fruit length
remove non-numeric trait values and check units
```{r}
FruitLength_mm_families_BIEN <- FruitLength_mm_families_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(FruitLength_mm_families_BIEN)
unique(FruitLength_mm_families_BIEN$unit)

FruitLength_mm_genus_BIEN <- FruitLength_mm_genus_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(FruitLength_mm_genus_BIEN)
unique(FruitLength_mm_genus_BIEN$unit)
```

### Seed mass
remove non-numeric trait values and check units
```{r}
SeedMass_g_families_BIEN <- SeedMass_g_families_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(SeedMass_g_families_BIEN)
unique(SeedMass_g_families_BIEN$unit)

SeedMass_g_genus_BIEN <- SeedMass_g_genus_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(SeedMass_g_genus_BIEN)
unique(SeedMass_g_genus_BIEN$unit)

@@

### Plant Height
remove non-numeric trait values and check units
```{r}
PlantHeight_m_families_BIEN <- PlantHeight_m_families_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(PlantHeight_m_families_BIEN)
unique(PlantHeight_m_families_BIEN$unit)

PlantHeight_m_genus_BIEN <- PlantHeight_m_genus_BIEN %>%
  filter(!is.na(as.numeric(trait_value))) %>%
  mutate(trait_value = as.numeric(trait_value))

nrow(PlantHeight_m_genus_BIEN)
unique(PlantHeight_m_genus_BIEN$unit)
```

## Combine new info with existing trait table

To fill in gaps of missing traits, use genus level data first then family level. Add column to trait dataframe of the level of trait measurement: species, genus, family.

### Subset dataframe to only keep required columns
```{r}
# for traits to family level from BIEN, keep columns of scrubbed_family, trait_name, trait_value
PlantLifespan_years_families_BIEN_df <- PlantLifespan_years_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
FruitType_families_BIEN_df <- FruitType_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
GrowthForm_families_BIEN_df <- GrowthForm_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
FruitLength_mm_families_BIEN_df <- FruitLength_mm_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
SeedMass_g_families_BIEN_df <- SeedMass_g_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
PlantHeight_m_families_BIEN_df <- PlantHeight_m_families_BIEN[, c("scrubbed_family", "trait_name", "trait_value")]
```

```{r}
# for traits to genus level from BIEN, keep columns of scrubbed_genus, trait_name, trait_value
PlantLifespan_years_genus_BIEN_df <- PlantLifespan_years_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
FruitType_genus_BIEN_df <- FruitType_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
GrowthForm_genus_BIEN_df <- GrowthForm_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
FruitLength_mm_genus_BIEN_df <- FruitLength_mm_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
SeedMass_g_genus_BIEN_df <- SeedMass_g_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
PlantHeight_m_genus_BIEN_df <- PlantHeight_m_genus_BIEN[, c("scrubbed_genus", "trait_name", "trait_value")]
```

### Get a single trait value for each family/genus
```{r}
# numeric traits
numeric_traits_combined <- function(df, level, traitname) {
  df$trait_value <- as.numeric(df$trait_value)
  if (level == "genus") {
    summary_data <- df %>%
      group_by(scrubbed_genus) %>%
          summarise(
            trait_value_mean = exp(mean(log(trait_value))),
            records_used = n(),
            variance = var(trait_value, na.rm = TRUE)
          )
      # Add trait name column
      summary_data$TraitName <- traitname

      # Rename the mean column to TraitValue
      summary_data <- summary_data %>%
        rename(trait_value = trait_value_mean)

    return(summary_data)
  } else if (level == "family") {
    summary_data <- df %>%
      group_by(scrubbed_family) %>%
      summarise(
            trait_value_mean = exp(mean(log(trait_value))),
            records_used = n(),
            variance = var(trait_value, na.rm = TRUE)
          )
      # Add trait name column
      summary_data$TraitName <- traitname

      # Rename the mean column to TraitValue
      summary_data <- summary_data %>%
        rename(TraitValue = trait_value_mean)

    return(summary_data)
  }
}

# non-numeric traits
cat_traits_combined <- function(df, level, traitname){
  if (level == "genus") {
    summary_data <- df %>%
      group_by(scrubbed_genus) %>%
      summarise(
        ModeValue = names(sort(table(trait_value), decreasing = TRUE))[1],
        records_used = n(),
        mode_freq = max(table(trait_value)),
        variance = 1 - (max(table(trait_value)) / n())
    )

    # Add trait name column
    summary_data$TraitName <- traitname

    # Rename the mean column to TraitValue
    summary_data <- summary_data %>%
      rename(TraitValue = ModeValue)

  return(summary_data)
  } else if (level == "family") {
    summary_data <- df %>%
      group_by(scrubbed_family) %>%
      summarise(
        ModeValue = names(sort(table(trait_value), decreasing = TRUE))[1],
        records_used = n(),
        mode_freq = max(table(trait_value)),
        variance = 1 - (max(table(trait_value)) / n())
    )

    # Add trait name column
    summary_data$TraitName <- traitname

    # Rename the mean column to TraitValue
    summary_data <- summary_data %>%
      rename(TraitValue = ModeValue)

  return(summary_data)
  }
}

```{r}
dim(PlantLifespan_years_families_BIEN_df)
PlantLifespan_years_families_BIEN_df_average <- numeric_traits_combined(PlantLifespan_years_families_BIEN_df, "family", "PlantLifespan_years")
dim(PlantLifespan_years_families_BIEN_df_average)

dim(PlantLifespan_years_genus_BIEN_df)
PlantLifespan_years_genus_BIEN_df_average <- numeric_traits_combined(PlantLifespan_years_genus_BIEN_df, "genus", "PlantLifespan_years")
dim(PlantLifespan_years_genus_BIEN_df_average)
```

```{r}
dim(FruitType_families_BIEN_df)
FruitType_families_BIEN_df_average <- cat_traits_combined(FruitType_families_BIEN_df, "family", "FruitType")
dim(FruitType_families_BIEN_df_average)

dim(FruitType_genus_BIEN_df)
FruitType_genus_BIEN_df_average <- cat_traits_combined(FruitType_genus_BIEN_df, "genus", "FruitType")
dim(FruitType_genus_BIEN_df_average)
```

```{r}
dim(GrowthForm_families_BIEN_df)
GrowthForm_families_BIEN_df_average <- cat_traits_combined(GrowthForm_families_BIEN_df, "family", "GrowthForm")
dim(GrowthForm_families_BIEN_df_average)

dim(GrowthForm_genus_BIEN_df)
GrowthForm_genus_BIEN_df_average <- cat_traits_combined(GrowthForm_genus_BIEN_df, "genus", "GrowthForm")
dim(GrowthForm_genus_BIEN_df_average)
```

```{r}
dim(FruitLength_mm_families_BIEN_df)
FruitLength_mm_families_BIEN_df_average <- numeric_traits_combined(FruitLength_mm_families_BIEN_df, "family", "FruitLength_mm")
dim(FruitLength_mm_families_BIEN_df_average)

dim(FruitLength_mm_genus_BIEN_df)
FruitLength_mm_genus_BIEN_df_average <- numeric_traits_combined(FruitLength_mm_genus_BIEN_df, "genus", "FruitLength_mm")
dim(FruitLength_mm_genus_BIEN_df_average)
```

```{r}
dim(SeedMass_g_families_BIEN_df)
SeedMass_g_families_BIEN_df_average <- numeric_traits_combined(SeedMass_g_families_BIEN_df, "family", "SeedMass_g")
dim(PlantLifespan_years_families_BIEN_df_average)

dim(SeedMass_g_genus_BIEN_df)
SeedMass_g_genus_BIEN_df_average <- numeric_traits_combined(SeedMass_g_genus_BIEN_df, "genus", "SeedMass_g")
dim(SeedMass_g_genus_BIEN_df_average)
```

```{r}
dim(PlantHeight_m_families_BIEN_df)
PlantHeight_m_families_BIEN_df_average <- numeric_traits_combined(PlantHeight_m_families_BIEN_df, "family", "PlantHeight_m")
dim(PlantHeight_m_families_BIEN_df_average)

dim(PlantHeight_m_genus_BIEN_df)
PlantHeight_m_genus_BIEN_df_average <- numeric_traits_combined(PlantHeight_m_genus_BIEN_df, "genus", "PlantHeight_m")
dim(PlantHeight_m_genus_BIEN_df_average)
```

### Add column of trait level (species, genus, family)
```{r}
# Add new column trait_level to long_plant_trait dataframe
long_plant_traits <- long_plant_traits %>%
  mutate(TraitLevel = ifelse(!is.na(TraitValue), "species", NA))
```
